Demographic Species Distribution Models

### 2019-10-11 manuscript/ms.Rmd
use Beverton-Holt carrying capacity model

# format parameter estimates
bbs_means <- bbs_draws_sry$summary$statistics[, "Mean"]
bbs_cis <- bbs_draws_sry$summary$quantiles[, c("2.5%", "97.5%")]
bbs_cov_names <- c("intercept", cache$get("all_covs"))
bbs_params <- cbind(data.frame(mean = bbs_means), bbs_cis)
bbs_params <- bbs_params[seq_len(2 * length(bbs_cov_names)), ]

@@

I specify an observation model for the presence-absence data in the breeding-bird survey:

\begin{align}
y_i &\sim Bernoulli(p_i) \\ \nonumber
p_i &= 1 - e^{-\Lambda_i} \\ \nonumber
\Lambda_i &= \alpha \lambda_i  \\ \nonumber
\lambda_i &= eigval(\mathbf{A_i})
\end{align}

where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$. Imake a statistically convenient but biologically naive assumption that the size of each population is proportional to its intrinsic growth rate - a relationship that would arise from a constant (not density-dependent) competitive pressure. This simplifying assumption has been made in other spatial population models such as dynamic range models [@pagelForecastingSpeciesRanges2012]. As discussed in the text below, DSDMs can easily incorporate more complex and realistic forms of density-dependence as required.
The parameter $\alpha$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey). This yields $\Lambda_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].

@@

\begin{align}
\beta_{F} &\sim N(\mu_F, \mathbf{\Sigma}_F) \\ \nonumber
\beta_{S} &\sim N(\mu_S, \mathbf{\Sigma}_S) \\ \nonumber
\gamma &\sim N(1.17, 2.10) \\ \nonumber
\alpha &\sim U(0, \infty)
\end{align}.

The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels. These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs. I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{Aj}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{Aj}) \sim N(1.12, 1.28)$, and $logit(S_{Jj}) \sim N(-0.05, 0.85)$.
These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed. The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$. The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{Aj})$: 1.12, and 1.28.
Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
In the absence of any meaningful prior information for the compound parameter $\alpha$, I assigned it an improper uniform prior on all positive values.

>>>>

# format parameter estimates
bbs_means <- bbs_draws_sry$summary$statistics[, "Mean"]
bbs_cis <- bbs_draws_sry$summary$quantiles[, c("2.5%", "97.5%")]
bbs_cov_names <- colnames(bbs_data$x_train)
bbs_params <- cbind(data.frame(mean = bbs_means), bbs_cis)
bbs_params <- bbs_params[seq_len(2 * length(bbs_cov_names)), ]

@@

I specify an observation model for presence-absence data:

\begin{align}
y_i &\sim Bernoulli(p_i) \\ \nonumber
p_i &= 1 - e^{-\psi \Lambda_i} \\ \nonumber
\Lambda_i &= ln(1 + e^{(\lambda_i - 1) M})  \\ \nonumber
\lambda_i &= eigval(\mathbf{A_i})
\end{align}

where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$. From the growth rate, the relative abundance of individuals $\Lambda_i$ is calculated from an expression that combined the carrying capacity expression from the Beverton-Holt model: $K_i = (\lambda_i - 1) M$, where $M$ quantifies how competition limits population sizes, transformed via the softplus function $ln(1 + e^{K_i})$ so that $\Lambda_i$ is positive for all parameter values. The softplus transformation is necessary since the Beverton-Holt expression would otherwise yield zero or negative relative abundances when $\lambda_i <= 1$), and can be thought of a smooth version of the Heaviside function, with $\Lambda_i$ almost identical to $K_i$ for most $lambda_i > 1$.
This formulation is statistically convenient, but as discussed in the text below, DSDMs can incorporate more complex and realistic forms of density-dependence if required.
The parameter $\psi$ is an individual-level detection rate so that $\psi \Lambda_i$ gives the expected number of individuals detected during the survey at observation $i$. This leads us to the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].

@@

\begin{align}
\beta_{F} &\sim N(\mu_F, \mathbf{\Sigma}_F) \\ \nonumber
\beta_{S} &\sim N(\mu_S, \mathbf{\Sigma}_S) \\ \nonumber
\gamma &\sim N(1.17, 2.10) \\ \nonumber
M &\sim U(0, \infty) \\ \nonumber
\psi &\sim U(0, 1)
\end{align}.

The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels. These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs. I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{Aj}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{Aj}) \sim N(1.12, 1.28)$, and $logit(S_{Jj}) \sim N(-0.05, 0.85)$.
These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed. The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$. The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{Aj})$: 1.12, and 1.28.
Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
In the absence of any meaningful prior information for the competition effect parameter $M$ or the detection rate parameter $\psi$, I assigned the former an improper uniform prior on all positive values, and the latter a uniform prior on the unit interval.
